PoVaTiM: A Possibilistic Valid-Time Model \\ \vspace*{0.04truein}

### 2012-04-15 ijcis-pos-model/sources/intro.tex
update witha new version

The concept of time itself is very complex to handle and interpret~\cite{klein94},~\cite{Shackle61}, though it is very natural and omnipresent. As information systems often attempt the modelling of natural objects, concepts or processes, they often require modelling temporal aspects or concepts. Thus, several proposals have been concerned with the obtaining of theoretical models that allow the modelling or representation of time~\cite{Bolour82},~\cite{Cru97}.

@@

A lot of research concerns temporal database models and their approaches to the modelling of time. The first efforts were towards the representation of historical information related to objects represented by records in a database~\cite{Clifford:1985:AHR:971699.318922}. Some proposals tried to extend the Entity Relationship Model~\cite{Klopprogge:1983}, without impact on any database standards like SQL~\cite{Sarda:1990:ESH:627277.627409}.

Notably, in 1994, ``A Consensus Glossary of Temporal Database Concepts'' was published~\cite{Dyreson1994}. For this publication, 44 temporal database researchers, among which some of the main researchers in this field, cooperated to reach a consensus on the nature and definitions of some of the main temporal database concepts and terminology. This glossary was subsequently updated in 1998~\cite{Jensen:Dyreson:1998:TheConsensusGlossary}.

An interesting issue in temporal modelling concerns relationships between temporal notions. Notably, Allen~\cite{Allen83} studied temporal relationships between time intervals (and as a special case time points). Among others, the querying of temporal databases has greatly profited from these temporal relationships, because they allowed for richer and more complex user-specified temporal query demands, by allowing to express more complex relationships between the temporal notions in the temporal expressions in the query and the temporal indications in the database. e.g., in a relation modelling who was department head of an institution during which periods of time, a query like `who were the department heads when Thomas worked for the institution' can be evaluated using similar relationships.

@@

To allow information systems to cope with these and similar data imperfections, many approaches adopt fuzzy sets~\cite{Zadeh65} for the representation of temporal information~\cite{343607},~\cite{nagypal2003},~\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy},~\cite{Dubois:jucs_9_9:fuzziness_and_uncertainty_in}. The temporal relationships studied by Allen were fuzzified by several authors~\cite{ohlbach2004},~\cite{nagypal2003},~\cite{schockaert08}. Garrido et al. ~\cite{garrido2009} present different temporal operators, defined by a combination of regular fuzzy comparisons. Both~\cite{garrido2009} and~\cite{Pon11} deal with uncertainty in temporal expressions concerning time intervals. Other approaches, like~\cite{Qia09}, use rough sets~\cite{Pawlak1995} to represent time intervals.

Next to temporal modelling, some attention has been spent on temporal reasoning~\cite{Allen83}. Although temporal reasoning is not discussed in this chapter, it should be noted that, among others, Dubois and Prade et al.~\cite{Dubois:jucs_9_9:fuzziness_and_uncertainty_in},~\cite{Dubois89} have dealt with fuzziness and uncertainty in temporal reasoning.

The aim of this chapter is to present and explain some main concepts regarding time in information systems and to present and discuss some issues and techniques concerned with handling data imperfections related to time. The rest of this chapter is structured as follows. Section \ref{sec:time-domain} presents some basic concepts and terminology about temporal modelling and discusses some of its important aspects and issues, while section~\ref{sec:fuzzy-time-mod} presents some important issues concerning the combination of data imperfections and temporal modelling. In section \ref{sec:temporal-databases}, some basic concepts and terminology about temporal databases are presented, followed by an overview of some interesting issues concerning temporal databases and a survey of some commercial temporal database systems. Finally, in section \ref{sec:imprecision-uncertainty}, an approach to querying temporal databases containing imperfect temporal information is presented, followed by some conclusions and some suggestions for future work in section \ref{sec:further-research-work}.

%The concept of \emph{time} is easy to understand but very complex to define~\cite{klein94}~\cite{Shackle61}. 
%In Information Systems \emph{IS}, several proposals have been concerned with the obtaining of theoretical models that allow the representation of time~\cite{Bolour82}~\cite{Cru97}. 
%More towards temporal databases, a lot of work have been done. The first efforts were towards the representation of the historical information related to an object in the database(reference). Some works tried to extend the Entity Relationship Model \emph{ERM} ~\cite{Klopprogge:1983} but without an impact on any database standards like SQL~\cite{Sarda:1990:ESH:627277.627409}.

%The ``Consensus Glossary of Temporal Database concepts'' \cite{Dyreson1994} is the first publication where the main researchers in temporal databases define the main thesaurus of temporal database concepts. This glossary defines the main types of time in a temporal database and is the basis of furthers researches ~\cite{Sarda:1990:ESH:627277.627409} \cite{Jensen94thetsql2}.

%Another main issue in temporal databases is the querying. The user specifies in the query some temporal constraints, e.g., \emph{``before this year''},\emph{``after 1995''}. In order to solve these temporal comparisons, between the time specified in the query and the temporal data stored in the database, some basic relationships have been defined. Allen~\cite{Allen83} first introduced the temporal relations between time intervals and, to a lesser extent, to time points. Some authors~\cite{ohlbach2004},\cite{nagypal2003},\cite{schockaert08} have softened the temporal operators defined by Allen. Therefore it is possible to compute the relationships between two ill-known time points or intervals. In~\cite{garrido2009}, some different temporal operators are defined by a combination of regular fuzzy comparisons. %e.g., the \emph{BEFORE} operator is defined by means of the fuzzy less than (FLT) operator.

%To allow information systems to cope with uncertainty, many approaches adopt fuzzy sets~\cite{Zadeh65} for the representation of temporal information~\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy},\cite{Dubois:jucs_9_9:fuzziness_and_uncertainty_in},~\cite{devos94}. The time points representing the time when the object is valid in the reality might not be precisely known. In order to deal with this, several models have been proposed. Garrido~\cite{garrido2009} proposes a model to deal with uncertainty in the time interval by means of fuzzy intervals. Bronselaer~\cite{Pon11} proposes a consistent framework to deal and represent ill-known time values and their relationships. Rough sets~\cite{Pawlak1995} have been also used to represent time intervals~\cite{Qia09} in databases.

%temporal reasoning and fuzzy temporal reasoning.
%Next to the manage of temporal information is the temporal reasoning~\cite{Allen83}. Dubois and Prade ~\cite{Dubois:jucs_9_9:fuzziness_and_uncertainty_in},\cite{Dubois89} deal with fuzziness and uncertainty in temporal reasoning, but this topic is outside the scope of this work.

%temporal commercial systems.

%The work is organized as follows:

%In Section \ref{sec:time-domain} an study about the time and its properties is done. Section \ref{sec:temporal-databases} is an overview of the main problems when managing the time in a database, the different proposals for solving them and the temporal database proposals. The section concludes with a comparison among the different proposals. In Section \ref{sec:fuzzy-temporal-databases} we present a sum up with the proposals for handling the imprecision in temporal databases and the main deficiencies we have detected. Finally, there are several commercial temporal database management systems  \emph{TDMBS} like \cite{oracle2009},\cite{posgree2009},\cite{teradata2011},\cite{timedb2005}. All of them are analyzed and compared later on this work.

>>>>

The concept of time itself is very complex to handle and interpret~\cite{Klein1994},~\cite{Shackle1961}, though it is very natural and omnipresent. As information systems often attempt the modelling of natural objects, concepts or processes, they often require modelling temporal aspects or concepts. Thus, several proposals have been concerned with the obtaining of theoretical models that allow the modelling or representation of time~\cite{Bolour1982},~\cite{VanderCruyssen1997}.

@@

A lot of research concerns temporal database models and their approaches to the modelling of time. The first efforts were towards the representation of historical information related to objects represented by records in a database~\cite{Clifford1985}. Some proposals tried to extend the Entity Relationship Model~\cite{Klopprogge1983}, without impact on any database standards like SQL~\cite{Sarda1990}.

Notably, in 1994, ``A Consensus Glossary of Temporal Database Concepts'' was published~\cite{Dyreson1994}. For this publication, 44 temporal database researchers, among which some of the main researchers in this field, cooperated to reach a consensus on the nature and definitions of some of the main temporal database concepts and terminology. This glossary was subsequently updated in 1998~\cite{Dyreson1998}.

An interesting issue in temporal modelling concerns relationships between temporal notions. Notably, Allen~\cite{Allen1983} studied temporal relationships between time intervals (and as a special case time points). Among others, the querying of temporal databases has greatly profited from these temporal relationships, because they allowed for richer and more complex user-specified temporal query demands, by allowing to express more complex relationships between the temporal notions in the temporal expressions in the query and the temporal indications in the database. e.g., in a relation modelling who was department head of an institution during which periods of time, a query like `who were the department heads when Thomas worked for the institution' can be evaluated using similar relationships.

@@

To allow information systems to cope with these and similar data imperfections, many approaches adopt fuzzy sets~\cite{Zadeh1965} for the representation of temporal information~\cite{Mitra1994},~\cite{Nagypal2003},~\cite{Billiet2011},~\cite{Dubois2003}. The temporal relationships studied by Allen were fuzzified by several authors~\cite{Ohlbach2004},~\cite{Nagypal2003},~\cite{Schockaert2008}. Garrido et al. ~\cite{Garrido2009} present different temporal operators, defined by a combination of regular fuzzy comparisons. Both~\cite{Garrido2009} and~\cite{Pons2011} deal with uncertainty in temporal expressions concerning time intervals. Other approaches, like~\cite{Qiang2009}, use rough sets~\cite{Pawlak1995} to represent time intervals.

Next to temporal modelling, some attention has been spent on temporal reasoning~\cite{Allen1983}. Although temporal reasoning is not discussed in this chapter, it should be noted that, among others, Dubois and Prade et al.~\cite{Dubois2003},~\cite{DuBois1989} have dealt with fuzziness and uncertainty in temporal reasoning.

The aim of this work is to present and explain the main operations of the manipulation language for a temporal database which stores the valid-time periods of the objects with uncertainty. The rest of the work is organized as follows. Section \ref{sec:prelim} presents some background concepts about both possibility theory and temporal databases. In section \ref{sec:time-rep} the representation of the valid-time intervals in the database is explained. Section \ref{sec:temporal-model} explains the main concepts and behaviour of the temporal model. The Data Manipulation Language (DML) is described and implemented. Finally in Section \ref{sec:conclusions} presents the conclusions and some possibilities for future research work.

### 2012-04-15 ijcis-pos-model/sources/temporaldatabases.tex
update witha new version

\begin{itemize}
	\item
	\textbf{Transaction time} (\emph{TT}) \cite{Rowe1987},\cite{Jensen1991} denotes the time when the fact (object) is stored in the database. It is usually append-only: as the past can not be changed, TT can not be changed neither. Furthermore, at the moment of insertion, a TT can be neither in the past nor in the future.
	\item
	\textbf{Valid time} (\emph{VT}) \cite{Jensen1994},\cite{Sarda1990},\cite{McKenzie1981} denotes the time when the fact (object) is true in the modelled reality. A fuzzy extension has been proposed by \cite{Garrido2009}. 
%	\item
%	User defined time: It is an uninterpreted attribute. The domain is date/time. The query language has no special support for it.
	\item
	\textbf{Decision time} (\emph{DT}, proposed in \cite{Nascimento1995}) denotes the time when an event was decided to happen. 
	\end{itemize}

@@

\subsubsection{Insert}
The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used.

There are three main cases when performing a create operation:
\begin{enumerate}
\item The object $A$ was never in the relation $R$: The object is added with the valid-time indicated by the crisp interval $I$.

\item The object $A$ was part of some previously recorded state: A new version of the object is inserted in the relation $R$ with the valid-time indicated by $I$.

\item The object $A$ is current in the relation $R$: The insert operation is rejected, a modify operation should be used.
\end{enumerate}

@@

\begin{align}
\label{eq:insert}
\INS \left(R ,A, I\right) =\\
\begin{cases}
\nonumber
R \cup \left(A, I \right), & \mbox{ if }  A \not \in R\\
R \cup  & \mbox{ if }  A \in R \wedge \\
R, & \mbox{ otherwise }  
\end{cases} 	
\end{align}

>>>>

\begin{itemize}
	\item
	\textbf{Transaction time} (\emph{TT}) \cite{Rowe1987},\cite{Jensen1991} denotes the time when the fact (object) is stored in the database. It is usually append-only: as the past can not be changed, TT can not be changed neither. Furthermore, at the moment of insertion, a TT can be neither in the past nor in the future.
	\item
	\textbf{Valid time} (\emph{VT}) \cite{Jensen1994},\cite{Sarda1990},\cite{McKenzie1981} denotes the time when the fact (object) is true in the modelled reality. A fuzzy extension has been proposed by \cite{Garrido2009}. 
%	\item
%	User defined time: It is an uninterpreted attribute. The domain is date/time. The query language has no special support for it.
	\item
	\textbf{Decision time} (\emph{DT}) \cite{Nascimento1995},\cite{Chakravarthy1994},\cite{Etzion1992},\cite{Ozsoyoglu1995} denotes the time when an event was decided to happen. 
	\end{itemize}

@@

In order to simplify the algorithms for the manipulation of data, some auxiliary functions are defined:

\begin{definition}
\label{def:close-a-crisp-interval-r}
Consider two crisp intervals $I= \left[a ,b \right]$ and $J= \left[c ,d \right]$, $c > a$. The CloseR$\left(I, J\right)$ function allows to close the right-open interval $I$ with respect to the first value $d$ in $J$:
\begin{align}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } b \neq +\infty \\
I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
\end{cases}
\end{align}
\end{definition}

Analogously, it is defined the function to close a left-open interval: 
\begin{definition}
\label{def:close-a-crisp-interval-l}
CloseL$\left(I, J\right)$ 
\begin{align}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } a \neq -\infty \\
I=\left]d, b \right] & \mbox{ if } a = -\infty \wedge J < I
\end{cases}
\end{align}
\end{definition}

\begin{definition}
\label{def:current-in-relation}
Consider an object $A$ and a relation $R$ in a temporal database. The function Current$\left(R, A \right)$ obtains the tuple $\left(A, I \right)$ that is current in the relation $R$, that is it, $I = \left[a, +\infty \right]$.

\begin{align}
\label{eq:current-in-relation}
\mbox{Current} \left(R, A \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } \exists b \in I : (A,I) \in R \wedge b = +\infty \\
\emptyset & \mbox{ in any other case }
\end{cases}
\end{align}
\end{definition}

\begin{definition}
\label{def:first-in-relation}
Analogously, the function First$\left(R, A \right)$ obtains the tuple $\left(A, I \right)$ where $I$ is a right-open interval and which is the first version in the relation $R$, that is it, $I = \left[-\infty, b \right]$.
\begin{align}
\label{eq:first-in-relation}
\mbox{First} \left(R, A \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } \exists b \in I : (A,I) \in R \wedge a = -\infty \\
\emptyset & \mbox{ in any other case }
\end{cases}
\end{align}

\end{definition}

Now it is possible to close the current version of an object by using \eqref{def:close-a-crisp-interval-r} and \eqref{eq:current-in-relation}

\begin{definition}
\label{def:close-current-version}
Consider an object $A$, the relation $R$ and a time interval $J$. Then, the function close-current$\left(R,A,J \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J \right)$.

\begin{eqnarray}
%\label{eq:close-current}
\text{Close-current} \left(R, A, J \right) =\\
\begin{cases}
\nonumber
\big \lbrace R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right) \cup \left(A, J\right)\right \rbrace  \big \rbrace \\
\nonumber
\mbox{ if } I = \mbox{ Current } \left(R, A \right) \wedge J > \mbox{ CloseR } \left(I, J \right)   \\
\nonumber R , \text{ in any other case}
\end{cases}
\end{eqnarray}
\end{definition}

It is also possible to append at the beggining by using \eqref{def:close-a-crisp-interval-l} and \eqref{def:first-in-relation}.
\begin{definition}
\label{def:append-first-version}
Consider an object $A$, the relation $R$ and a time interval $J$. Then, the function append-first$\left(R,A,J \right)$ closes any right open version of the object $A$ if it exists and add the new version $\left(A, J \right)$.

\begin{eqnarray}
%\label{eq:close-current}
\text{Append-first} \left(R, A, J \right) =\\
\begin{cases}
\nonumber
\big \lbrace R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseL } \left(I, J\right) \right) \cup \left(A, J\right)\right \rbrace  \big \rbrace \\
\nonumber
\mbox{ if } I = \mbox{ First } \left(R, A \right) \wedge J < \mbox{ CloseL } \left(I, J \right)   \\
\nonumber R , \text{ in any other case}
\end{cases}
\end{eqnarray}
\end{definition}

\subsubsection{Insert}
The user wants to store an object $A$ which is valid in the relation $R$ during the time interval specified by $I = \left(a, b \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are two main cases when performing a create operation:
\begin{enumerate}
\item The object $A$ was never in the relation $R$: The object is added with the valid-time indicated by the crisp interval $I$.

\item The object $A$ is in the relation $R$. Depending on the value of the time interval $I$, there are two possibilities:
	\begin{enumerate}
	\item Insert $\left(A, I\right)$ in the relation $R$. If the time interval $I$ does not overlap any other valid-time interval in the relation $R$.
	\item Reject the insertion, if the time interval $I$ do overlap any existing valid-time interval for the object $A$ in the relation $R$.
	\end{enumerate}

\end{enumerate}

@@

\begin{align}
\label{eq:insert}
\INS \left(R ,A, I\right) =\\
\begin{cases}
\nonumber
R \cup \left(A, I \right), & \mbox{ if }  A \not \in R\\
R \cup   \left(A, I \right) & \mbox{ if }  A \in R \wedge \\
& \big \lbrace \left( \exists J: \left(A, J \right) \in R \wedge J<I  \right) \vee \\
& \left( \exists K: \left(A, K \right) \in R \wedge K>I \vee \right) \wedge\\
&  J<I<K  \big \rbrace\\
R, & \mbox{ otherwise }  
\end{cases} 	
\end{align}
